Compositional Quantitative Resource Analysis of C Programs

### 2014-06-18 paper/main.tex
switch to smallstep

To formally describe the cost of operation of a given
program, our semantics tracks resource consumption while
executing.  The semantics is oblivious to the resource of
interest for the user of our system.  We acheive this
independence by parameterizing the semantics by a resource
\emph{metric}, it is a tuple of rational numbers with the
following form
$$
(
M_a, M_s, M_b, M_r, M_u, M_f,
M_l,
M_c^1, M_c^2,
M_q
).
$$
Each of these rational numbers indicates the amount of resource
consumed by an operation.  If one of these numbers is negative,
it means that resources are released by the program.

@@

\Rule{S:Abort}
{ }
{ (\state, S) \Downarrow_{(0,0)} \circ }

\and \Rule{S:Assert}
{ {\rm istrue}~\llbracket e \rrbracket_\state}
{ (\state, \code{assert}~e) \Downarrow_{M_a} \code{seq}~\state }

\and \Rule{S:Skip}
{ }
{ (\state, \code{skip}) \Downarrow_{M_s} \code{seq}~\state }

\and \Rule{S:Break}
{ }
{ (\state, \code{break}) \Downarrow_{M_b} \code{brk}~\state }

\and \Rule{S:Return}
{ }
{ (\state, \code{return}~e) \Downarrow_{M_r}
  \code{ret}~\state~\llbracket e \rrbracket_\state
}

\and \Rule{S:Update}
{ }
{ (\state, x \gets e) \Downarrow_{M_u}
  \code{seq}~\state[x \mapsto \llbracket e \rrbracket_\state]
}

\and \Rule{S:Call}
{ \delta~f = (\vec y, S) \\
  ((\state, y_i \mapsto \state(x_i)), S)
    \Downarrow_{(q,q')} \code{ret}~\state'~v
}
{ (\state, r \gets f(\vec x)) \Downarrow_{M_f\cdot(q,q')}
  \code{seq}~(\state'~\backslash~(x_i))[r \mapsto v]
}

\and \Rule{S:LoopSeq}
{ (\state, S) \Downarrow_{(q_1, q_1')} \code{seq}~\state'
\\ (\state', \code{loop}~S) \Downarrow_{(q_2, q_2')} O }
{ (\state, \code{loop}~S) \Downarrow_{(q_1, q_1') \cdot M_l \cdot (q_2, q_2')} O }

\and \Rule{S:LoopBrk}
{ (\state, S) \Downarrow_{(q, q')} \code{brk}~\state' }
{ (\state, \code{loop}~S) \Downarrow_{(q, q')} \code{seq}~\state' }

\and \Rule{S:LoopAbort}
{ (\state, S) \Downarrow_{(q, q')} O
\\ \forall \state.~O \neq \code{seq}~\state \land O \neq \code{brk}~\state }
{ (\state, \code{loop}~S) \Downarrow_{(q, q')} O }

\and \Rule{S:IfThen}
{  (\state, S_1) \Downarrow_{(q, q')} O
\\ {\sf istrue}~\llbracket e \rrbracket_\state
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2)
  \Downarrow_{M_c^1 \cdot (q, q')} O
}

\and \Rule{S:IfElse}
{  (\state, S_2) \Downarrow_{(q, q')} O
\\ {\sf isfalse}~\llbracket e \rrbracket_\state
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2)
  \Downarrow_{M_c^2 \cdot (q, q')} O
}

\and \Rule{S:Seq}
{  (\state, S_1) \Downarrow_{(q_1, q'_1)} \code{seq}~\state'
\\ (\state', S_2) \Downarrow_{(q_2, q'_2)} O
}
{ (\state, S_1; S_2) \Downarrow_{(q_1, q'_1) \cdot M_q \cdot (q_2, q'_2)} O }

\and \Rule{S:SeqAbort}
{ (\state, S_1) \Downarrow_{(q,q')} O
\\ \forall \state.~O \neq \code{seq}~\state}
{ (\state, S_1; S_2) \Downarrow_{(q,q')} O }

@@

In the evaluation judgement $(\state, S) \Downarrow_{(q, q')} O$,
the outcome $O$ specifies how the statement $S$ terminated.
$$
O := \circ
\mid \code{seq}~\sigma
\mid \code{brk}~\sigma
\mid \code{ret}~\sigma~v
$$
The \code{seq} outcome is the simple fallthrough outcome, for example
a \code{skip} statement will terminate with this outcome without changing
the heap as pictured in the rule {\sc S:Skip}.  The \code{brk} and
\code{ret} outcomes describe a statement that terminated with a break or
return statement.  In the latter case, a return value is also part of the
outcome.  The use of these different outcomes can be explained by considering
the loop statement rules:  If the inner statement breaks
then the loop exits using the rule {\sc S:LoopBrk}, if the inner statement
falls through another iteration is run using the rule {\sc S:LoopSeq},
and all other cases are handled in the rule {\sc S:LoopAbort}.

@@

The busy result is also useful to handle diverging programs with big-step
judgements.  In case a program does not terminate, it is impossible to
derive $(\state, S)\Downarrow_{(q, q')} O$ with $O \neq \circ$.  The only
possible way to conclude a big-step judgement is to use the {\sc S:Abort}
rule at some point in the computation.  One alternative to this technique
is to use coinductive semantics~\cite{Leroy-coind}, however, using the busy
result prevents duplication and allows to have only one soundness proof for
both unsafe and non-terminating programs.  Indeed, stating a result for
$(\sigma, S) \Downarrow_{(q,q')} \circ$ covers unsafe programs, diverging
programs, and partial computations at the same time.

@@

\begin{figure*}[t]
\begin{mathpar}

\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q \implies \code{istrue}~\sem{e}_\state }
  { \code{assert}~e }
  { Q }
}

\and
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_s + Q }{ \code{skip} }{ Q }
}

\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}

\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_r + R\,\sem{e}_\state }{ \code{return}~e }{ Q }
}

\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}

\and
\Rule{L:Call}
{
  \Delta f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z) \\
  P \models P_f \, y \, (\sigma(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \sigma[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q }
}

\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}

\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}

\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { P \land \code{istrue}~\sem{e}_\state - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { P \land \code{isfalse}~\sem{e}_\state - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}

\and
\Rule{L:Weaken}
{
  P \models P' \\
  \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \\
  Q' \models Q \\
  B' \models B \\
  \forall v.\, (R'\,v \models R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}

\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \mathbb Q_0^+
}
{ \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x }
}

\and
\Rule{L:Extend}
{ \forall f\,P_f\,Q_f.\,
  \Delta'f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ \delta f }{ \bot })
\\
  \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}

\end{mathpar}

>>>>

To formally describe the cost of operation of a given
program, our semantics tracks resource consumption while
executing.  The semantics is oblivious to the resource of
interest for the user of our system.  We acheive this
independence by parameterizing the semantics by a resource
\emph{metric}, it is a tuple of rational numbers with the
following form
$$
(
M_a, M_b, M_r, M_u, M_f,
M_l,
M_c^1, M_c^2,
M_q, XXX
).
$$
Each of these rational numbers indicates the amount of resource
consumed by an operation.  If one of these numbers is negative,
it means that resources are released by the program.

@@

\begin{figure}[ht!]
\begin{mathpar}
%
\Rule{S:Assert}
{ \code{istrue}~\sem{e}_\state }
{ (\state, \code{assert}~e, \cont, \cost)
\smallstep
(\state, \code{skip}, \cont, \cost {-} M_a)
}
%
\and
\Rule{S:BrkSeq}
{}
{ (\state, \code{break}, \Kseq S \cont, \cost)
\smallstep
(\state, \code{break}, \cont, \cost)
}
%
\and
\Rule{S:BrkLoop}
{}
{ (\state, \code{break}, \Kloop \cont, \cost)
\smallstep
(\state, \code{skip}, \cont, \cost {-} M_b)
}
%
\and
\Rule{S:RetSeq}
{}
{ (\state, \code{return}~e, \Kseq S \cont, \cost)
\smallstep
(\state, \code{return}~e, \cont, \cost)
}
\and
%
\and
\Rule{S:RetLoop}
{}
{ (\state, \code{return}~e, \Kloop \cont, \cost)
\smallstep
(\state, \code{return}~e, \cont, \cost)
}
%
\and
\Rule{S:RetCall}
{ \state = (\_, \gamma) \\
  \state' = (\theta, \gamma)[r \mapsto \state(x)]
}
{ (\state, \code{return}~e, \Kcall r \theta \cont, \cost)
\smallstep
(\state', \code{skip}, \cont, \cost {-} M_r)
}
%
\and
\Rule{S:Update}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\smallstep
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
%
\and
\Rule{S:Call}
{ \delta f = (\vec x, S_f) \\
  \state = (\theta, \gamma) \\
  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
}
{ (\state, r \gets f(\vec y), \cont, \cost)
\smallstep
(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
}
%
\and
\Rule{S:Loop}
{}
{ (\state, \code{loop}~S, \cont, \cost)
\smallstep
(\state, S, \Kloop S \cont, \cost)
}
%
\and
\Rule{S:SkipLoop}
{}
{ (\state, \code{skip}, \Kloop S \cont, \cost)
\smallstep
(\state, S, \Kloop S \cont, \cost {-} M_l)
}
%
\and
\Rule{S:IfTrue}
{ \code{istrue}~\sem{e}_\sigma }
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep
(\state, S_1, \cont, \cost {-} M_c^1)
}
%
\and
\Rule{S:IfFalse}
{ \code{isfalse}~\sem{e}_\sigma }
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep
(\state, S_2, \cont, \cost {-} M_c^2)
}
%
\and
\Rule{S:Seq}
{ }
{ (\state, S_1; S_2, \cont, \cost)
\smallstep
(\state, S_1, \Kseq {S_2} \cont, \cost)
}
%
\and
\Rule{S:SkipSeq}
{}
{ (\state, \code{skip}, \Kseq S \cont, \cost)
\smallstep
(\state, S, \cont, \cost {-} M_q)
}
%

@@

In the evaluation judgement $(\state, S) \Downarrow_{(q, q')} O$,
the outcome $O$ specifies how the statement $S$ terminated.
$$
O := \circ
\mid \code{seq}~\state
\mid \code{brk}~\state
\mid \code{ret}~\state~v
$$
The \code{seq} outcome is the simple fallthrough outcome, for example
a \code{skip} statement will terminate with this outcome without changing
the heap as pictured in the rule {\sc S:Skip}.  The \code{brk} and
\code{ret} outcomes describe a statement that terminated with a break or
return statement.  In the latter case, a return value is also part of the
outcome.  The use of these different outcomes can be explained by considering
the loop statement rules:  If the inner statement breaks
then the loop exits using the rule {\sc S:LoopBrk}, if the inner statement
falls through another iteration is run using the rule {\sc S:LoopSeq},
and all other cases are handled in the rule {\sc S:LoopAbort}.

@@

The busy result is also useful to handle diverging programs with big-step
judgements.  In case a program does not terminate, it is impossible to
derive $(\state, S)\Downarrow_{(q, q')} O$ with $O \neq \circ$.  The only
possible way to conclude a big-step judgement is to use the {\sc S:Abort}
rule at some point in the computation.  One alternative to this technique
is to use coinductive semantics~\cite{Leroy-coind}, however, using the busy
result prevents duplication and allows to have only one soundness proof for
both unsafe and non-terminating programs.  Indeed, stating a result for
$(\state, S) \Downarrow_{(q,q')} \circ$ covers unsafe programs, diverging
programs, and partial computations at the same time.

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q \implies \code{istrue}~\sem{e}_\state }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_s + Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_r + R\,\sem{e}_\state }{ \code{return}~e }{ Q }
}
%
\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z) \\
  P \models P_f \, y \, (\state(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { P \land \code{istrue}~\sem{e}_\state - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { P \land \code{isfalse}~\sem{e}_\state - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Weaken}
{
  P \models P' \\
  \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \\
  Q' \models Q \\
  B' \models B \\
  \forall v.\, (R'\,v \models R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \mathbb Q_0^+
}
{ \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x }
}
%
\and
\Rule{L:Extend}
{ \forall f\,P_f\,Q_f.\,
  \Delta'f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ \delta f }{ \bot })
\\
  \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\end{mathpar}
